---
title: "ULAB"
author: "Austin"
date: "03/09/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyr)
library(nlme)
library(ggplot2)
library(dplyr)
#install.packages("lubridate") you will need to run this line the first time you use a package, comment it out after.
library(lubridate)
#install.packages("zoo")
library(zoo)
#install.packages("nlme")
library(nlme)
```

## R Markdown

```{r}
# load data from csv into R dataframe
datapath <- file.path('/Users','austinle','Downloads') # point to directory with data file (NEED TO CHANGE)
data_fname <- file.path(datapath,"Subjects_TP (1).csv") # path to spreadsheet
wideformda <- read.csv(data_fname) # wide form since each visit has its own column, one row per individual
colnames(wideformda)[colnames(wideformda)=="Diag.AV45"] <- "Diag.AV45_1" # rename one annoying column
wideformda$baseline_AV45_TP_CorticalSummary.CerebellarWhole_1 <- wideformda$AV45_NONTP_CorticalSummary.CerebellarWhole_1
wideformda$baseline_AV45_TP_CorticalSummary.BigRef_1 <- wideformda$AV45_NONTP_CorticalSummary.BigRef_1
colnames(wideformda)[colnames(wideformda)=="AV45_NONTP_CorticalSummary.CerebellarWhole_1"] <- "AV45_TP_CorticalSummary.CerebellarWhole_1"
colnames(wideformda)[colnames(wideformda)=="AV45_NONTP_CorticalSummary.BigRef_1"] <- "AV45_TP_CorticalSummary.BigRef_1"

wideformda
```


```{r}
# vector of all variable names we want to include in our analyses (NEED TO ADD COLUMNS)
desired_column_names <- c("AV45_Date",
                          "Age.AV45",
                          "Diag.AV45",
                          "Diag.AV45_1",
                          "UW_MEM_AV45",
                          "UW_EF_AV45",
                          "AV45_TP_CorticalSummary.CerebellarWhole",
                          "baseline_AV45_TP_CorticalSummary.CerebellarWhole_1",
                          "AV45_TP_CorticalSummary.CerebellarWhole_2",
                          "AV45_TP_CorticalSummary.BigRef",
                          "baseline_AV45_TP_CorticalSummary.BigRef_1",
                          "AV45_TP_CorticalSummary.BigRef_2")
                          #"UW_TEST_EXAMDATE",
                          #"UW_EF",
                          #"UW_MEM")

# vector of variable names that don't change from visit to visit
baseline_vars <- c("SID","Gender","APOE2_BIN","APOE4_NUM","Edu..Yrs.")
```

```{r}
longformda <- list() # initialize empty list
# loop through each desired column above and create a list of data frames in long format with only the baseline variables and data from each visit for that column
for (colname in desired_column_names) {
  # all column beginning with the desired column name
  columns_to_reshape <- c(names(wideformda)[startsWith(names(wideformda), colname)])
  longformda[[colname]] <- reshape( # wide format data, baseline variables and desired columns only  
    data=wideformda[c(baseline_vars,columns_to_reshape)],
    # specify that we're converting from wide to long format
    direction="long",
    # specify that we want to reshape these desired columns
    varying=columns_to_reshape,
    # define a new variable that will track what visit number the data comes from
    timevar="visitnumber",
    # indicate number of different visits with data for this variable
    times=c(as.character(1:length(columns_to_reshape))),
    # we name the column the desired column name
    v.names=colname,
    # indicate which variable we use to identify individuals
    idvar="SID")
}
```

```{r}
# now that we have a list of data frames, this code merges all the data frames together in a compact way
finalda <- Reduce(
  function(x,y,...) merge(x, y, by=c(baseline_vars,"visitnumber"), all=TRUE), # merge "by" all the baseline variables and visit number
  longformda)
# order data by subject ID and visit number
finalda$visitnumber <- as.integer(finalda$visitnumber)
finalda <- finalda[order(finalda$SID, finalda$visitnumber),]
finalda
```

# save formatted data frame to csv file for easy viewing
#write.csv(finalda, file.path(datapath,"longform_spreadsheet.csv"), row.names=FALSE)

```{r}
write.csv(finalda, file.path(datapath,"longform_spreadsheet.csv"), row.names=FALSE)
```


Defining our sample
```{r}
# subset to only cognitively normal individuals (Diag.AV45=="N") and to only individuals with 2+ PET scans (if AV45_TP_CorticalSummary/CerebellarWhole_2 is NA, remove)

newdata <- filter(finalda, Diag.AV45_1 == "N", AV45_TP_CorticalSummary.BigRef_2 != "NA", AV45_TP_CorticalSummary.CerebellarWhole_2 != "NA") 

newdata

nrow(newdata) #Gives us the number of cognitively normal subjects at baseline scan 241

finaldata <- inner_join(select(newdata, SID), finalda, by= "SID")
finaldata

```

Replicating Table 1 from the paper (on baseline amyloid negative)
```{r}
# compute summary statistics (see Table 1, don't need to include every single variable)
finaldata_abnegative <- finaldata %>% filter(baseline_amyloid_status=="AB-")

#Total (Value in Table 1: 142)
Total = length(unique(finaldata_abnegative$SID)) 
#157

#Florbetapir follow up time, y? (Value in Table 1: 3.9 ± 1.4)

#Age at baseline, y (Value in Table 1: 74.7 ± 7.0)
FirstVisit = filter(finaldata_abnegative, visitnumber == 1) #filter for all first visits
AgeatBaseline = mean(FirstVisit$Age.AV45) #find mean age at first visit
#74.2160
AgeatBaselineSD = sd(FirstVisit$Age.AV45) #find st dev for age at first visit
#6.98805

#Female (Values in Table 1: 72 and 45%)
FilteredforFemale = filter(finaldata_abnegative, Gender == 2) #filter for all females
Female = length(unique(FilteredforFemale$SID)) #find number of unique female SIDs
#68
FemalePercent = Female / length(unique(finaldata_abnegative$SID)) #number of females divided by number of participants
#43.31 percent

#APOE4+ (Values in Table 1: 32 and 15%)
APOE4Status = filter(finaldata_abnegative, APOE4_NUM == 1) #filtered for all APOE4+
APOE4Plus = length(unique(APOE4Status$SID)) #find number of APOE4+
#30

APOE4PlusPercent = APOE4Plus / length(unique(finaldata_abnegative$SID)) #number of APOE4+ divided by number of participants
#19.11 percent

#Education, y (Value in Table 1: 16.7 ± 2.7)
Education = mean(FirstVisit$Edu..Yrs.) #find mean of education years at first visit
#16.656
EducationSD = sd(FirstVisit$Edu..Yrs.) #find st dev for education at first visit
#2.6691
#Converted to MCI/AD (Values in Table 1: MCI = 12, AD = 2, Percent = 9.9%)
MCI = filter(finaldata_abnegative, visitnumber != 1, Diag.AV45 == "LMCI") #find anyone MCI at any point
ConvertedtoMCI = length(unique(MCI$SID)) #number of people who converted to MCI because had filtered to cognitively normal at baseline
#18
AD = filter(finaldata_abnegative, visitnumber != 1, Diag.AV45 == "AD") #find anyone AD at any point
ConvertedtoAD = length(unique(AD$SID)) #number of people who converted to AD because had filtered to cognitively normal at baseline
#2
ConvertedtoMCIorADPercent = (ConvertedtoMCI + ConvertedtoAD)/length(unique(finaldata_abnegative$SID)) #number of people converted to MCI or AD divided by number of participants
#12.738 percent

#Converted to florbetapir+ (Values in table 1: 13 and 9.2%)

#Memory Composite (Value in Table 1: 1.12 ± 0.58)
Memory = mean(FirstVisit$UW_MEM_AV45) #find mean of memory score at first visit
#1.1355
MemorySD = sd(FirstVisit$UW_MEM_AV45) #find st dev for memory score at first visit
#0.5637

#Executive Function Compostie (Value in Table 1: 1.00 ± 0.67))
ExecFunction = mean(FirstVisit$UW_EF_AV45) #find mean of exec function score at first visit
#1.12
ExecFunctionSD = sd(FirstVisit$UW_EF_AV45) #find st dev for exec function score at first visit
#0.753

#Cognitive Testing Follow-up, y (Value in Table 1: 3.8 ± 1.0)


```


Computing other variables needed for analysis
```{r}
# compute baseline amyloid status (if AV45_NONTP_CorticalSummary/CerebellarWhole_1 > 1.11 (see paper), then AB+)

baseline_amyloid_status <- c()

for (i in 1:length(newdata$SID))
  if (newdata$baseline_AV45_TP_CorticalSummary.BigRef_1[i] < 0.79 &&
      newdata$baseline_AV45_TP_CorticalSummary.CerebellarWhole_1[i] < 1.11) {
      baseline_amyloid_status <- baseline_amyloid_status %>% append("AB-")
  }else if (newdata$baseline_AV45_TP_CorticalSummary.BigRef_1[i] > 0.79 &&
      newdata$baseline_AV45_TP_CorticalSummary.CerebellarWhole_1[i] > 1.11) {
      baseline_amyloid_status <- baseline_amyloid_status %>% append("AB+")
  } else {
      baseline_amyloid_status <- baseline_amyloid_status %>% append("discordant") 
  }

baseline_amyloid_status
table(baseline_amyloid_status) #241 participants, 157 AB-, 61 AB+, 23 discordant 

newdata$baseline_amyloid_status <- baseline_amyloid_status

finaldata <- inner_join(select(newdata, SID, baseline_amyloid_status), finalda, by= "SID")
finaldata

```

We still need to compute the time variable:
```{r}
# compute "time" variable: amount of time (days, months, years, w/e) in between AV45 scan and cognitive test (use dates of each from spreadsheet)
finaldata$AV45_Date <- mdy(finaldata$AV45_Date) # convert date strings to datetime using lubridate package
finaldata$time <- as.data.frame(finaldata %>% group_by(SID) %>% mutate(time = (AV45_Date - AV45_Date[1])/365))$time # create new variable "time" in years between first AV45 visit and current visit

finaldata
```

Compute AV45 slope
```{r}
# compute AV45 slope using linear regression (could be tricky, lmk if you need help:
# 1) run lm(AV45_TP_CorticalSummary... ~ Age) for each individual
# 2) the resulting "Estimate" value for Age is the AV45 slope for that individual

finaldata$annualized_AV45 <- as.data.frame(finaldata %>% group_by(SID) %>% # for each individual
  # regress AV45 SUVR against time (more accurate than Age), extract estimate for time (slope)
  mutate(annualized_AV45 = lm(AV45_TP_CorticalSummary.BigRef ~ time)$coefficients["time"]))$annualized_AV45

finaldata$annualized_AV45
```

Spaghetti plots!
```{r}
# figure 1 from paper
# use ggplot to plot age vs. AV45 SUVR for AB-

baseline_negative <- finaldata %>% select(SID, Age.AV45, AV45_TP_CorticalSummary.BigRef, baseline_amyloid_status) %>% filter(baseline_amyloid_status == "AB-") %>% na.omit

baseline_negative$SID <- as.factor(baseline_negative$SID) #We want SID to be categorical variable, otherwise R thinks it's a numerical value and will assign colors based on gradient

p <- ggplot(data = baseline_negative, aes(x = Age.AV45, y = AV45_TP_CorticalSummary.BigRef, group = SID)) +geom_point(aes(color = SID)) + geom_line(aes(color = SID)) + geom_hline(yintercept = 0.79, linetype = "dotted") + theme(legend.position = "none") +xlim(50, 100)+ylim(0.65, 0.95) + ggtitle("Baseline Negative") + labs(y="Florbetapir SUVR", x = "Age (years)")+
  theme(plot.title = element_text(hjust = 0.5))

p

# use ggplot to plot age vs. AV45 SUVR for AB+

baseline_positive <- finaldata %>% select(SID, Age.AV45, AV45_TP_CorticalSummary.BigRef, baseline_amyloid_status) %>% filter(baseline_amyloid_status == "AB+") %>% na.omit

baseline_positive$SID <- as.factor(baseline_positive$SID) #We want SID to be categorical variable, otherwise R thinks it's a numerical value and will assign colors based on gradient

p2 <- ggplot(data = baseline_positive, aes(x = Age.AV45, y = AV45_TP_CorticalSummary.BigRef, group = SID)) +geom_point(aes(color = SID)) + geom_line(aes(color = SID)) + geom_hline(yintercept = 0.79, linetype = "dotted") + theme(legend.position = "none") +xlim(50, 100)+ylim(0.70, 1.25) + ggtitle("Baseline Positive") + labs(y="Florbetapir SUVR", x = "Age (years)")+
  theme(plot.title = element_text(hjust = 0.5))

p2

```


Creating a new dataset for use in linear mixed effects modeling
```{r}
#using finaldata works in the other analyses, but for LME we will need to remove the NA values. Participant with SID 6051 has NA values for APOE2 and APOE4 statuses, so will be removed from LME modeling. 240 subjects used in LME modeling as opposed to 241 subjects in prior analyses.

finaldata

finaldata2 <- finaldata
finaldata2$Diag.AV45_1 <- na.locf(finaldata2$Diag.AV45_1)
finaldata2$baseline_AV45_TP_CorticalSummary.BigRef_1 <- na.locf(finaldata2$baseline_AV45_TP_CorticalSummary.BigRef_1)

finaldata2 <- finaldata2 %>% select(-baseline_AV45_TP_CorticalSummary.CerebellarWhole_1, -AV45_TP_CorticalSummary.CerebellarWhole_2, -AV45_TP_CorticalSummary.BigRef_2)

finaldata2 <- na.omit(finaldata2) #finaldata2 now has no NA values

```
Linear mixed effects model (LME) on baseline AB negative finaldata2

```{r}
# look into nlme package, same basic structure as lm function: outcome ~ independent variables, also need to define:
finaldata2_abnegative <- finaldata2 %>% filter(baseline_amyloid_status=="AB-")

#ModelA: Longitudinal memory as dependent variable

#at least 2 scans
summary(lme((UW_MEM_AV45 ~ Gender + APOE4_NUM + Edu..Yrs. + Age.AV45 + time + baseline_AV45_TP_CorticalSummary.BigRef_1 + time * baseline_AV45_TP_CorticalSummary.BigRef_1 + annualized_AV45 + time * annualized_AV45), data = finaldata2_abnegative, random = ~1+time|SID, control = lmeControl(opt="optim",maxIter=100,msMaxIter=100,niterEM=50,msMaxEval=400) ))

#ModelB: Longitudinal executive function as dependent variable

#at least 2 scans
summary(lme((UW_EF_AV45 ~ Gender + APOE4_NUM + Edu..Yrs. + Age.AV45 + time + baseline_AV45_TP_CorticalSummary.BigRef_1 + time * baseline_AV45_TP_CorticalSummary.BigRef_1 + annualized_AV45 + time * annualized_AV45), data = finaldata2_abnegative, random = ~1+time|SID, control = lmeControl(opt="optim",maxIter=100,msMaxIter=100,niterEM=50,msMaxEval=400) ))

# random effects (intercept and slope coded as "~1+time|SID") this is treating time as a variable so we can how the change in time relates to changes in time and amyloid, interaction between time and amyloid and how that affects cognition 

# model control parameters (how hard your model tries to "converge", you can google this if interested but I would use these very liberal parameters: control = lmeControl(opt="optim",maxIter=100,msMaxIter=100,niterEM=50,msMaxEval=400))
```

Subsetting finaldata2 based on number of scans
```{r}
length(unique((finaldata2_abnegative %>%
  group_by(SID) %>%
  filter(n() == 2))$SID))
#42 had exactly 2 scans
length(unique((finaldata2_abnegative %>%
  group_by(SID) %>%
  filter(n() == 3))$SID))
#44 had exaclty 3 scans
length(unique((finaldata2_abnegative %>%
  group_by(SID) %>%
  filter(n() == 4))$SID))
#41 had exactly 4 scans
length(unique((finaldata2_abnegative %>%
  group_by(SID) %>%
  filter(n() == 5))$SID))
#28 had exactly 5 scans
length(unique((finaldata2_abnegative %>%
  group_by(SID) %>%
  filter(n() == 6))$SID))
#1 had exactly 6 scans

length(unique((finaldata2_abnegative %>%
  group_by(SID) %>%
  filter(n() >2))$SID))
#114 have at least 3 scans

length(unique((finaldata2_abnegative %>%
  group_by(SID) %>%
  filter(n() >1))$SID))
#156 have at least 2 scans

```

LME model at least 3 scans
```{r}
finaldata2_abnegative_atleast3scans <- finaldata2_abnegative %>%
  group_by(SID) %>%
  filter(n() >2)

#modelA: Longitudinal memory as dependent variable, at least 3 scans
summary(lme((UW_MEM_AV45 ~ Gender + APOE4_NUM + Edu..Yrs. + Age.AV45 + time + baseline_AV45_TP_CorticalSummary.BigRef_1 + time * baseline_AV45_TP_CorticalSummary.BigRef_1 + annualized_AV45 + time * annualized_AV45), data = finaldata2_abnegative_atleast3scans, random = ~1+time|SID, control = lmeControl(opt="optim",maxIter=100,msMaxIter=100,niterEM=50,msMaxEval=400) ))

#modelB: Longitudinal executive function as dependent variable, at least 3 scans
summary(lme((UW_EF_AV45 ~ Gender + APOE4_NUM + Edu..Yrs. + Age.AV45 + time + baseline_AV45_TP_CorticalSummary.BigRef_1 + time * baseline_AV45_TP_CorticalSummary.BigRef_1 + annualized_AV45 + time * annualized_AV45), data = finaldata2_abnegative_atleast3scans, random = ~1+time|SID, control = lmeControl(opt="optim",maxIter=100,msMaxIter=100,niterEM=50,msMaxEval=400) ))
```

Replicating Figure 3 from paper (if we have time)

```{r}
# if time and interesting in visualization
# ggplot to make scatter plot of memory slope vs. annualized AV45 change, fit with geom_smooth(method='lm')

```


