---
title: "ULAB_AD_Mina"
author: "Mina Baghai"
date: "2/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

``` {r}
# load data from csv into R dataframe
datapath <- file.path('/Users', 'minabaghai', 'OneDrive', 'Documents', 'Berkeley', 'ULab') # point to directory with data file
data_frame <- file.path(datapath, "Subjects_TP1.csv") # Path to spreadsheet 
wideformda <- read.csv(data_frame) # wide form since each visit has its own column, one row per individual
colnames(wideformda)[colnames(wideformda)=="Diag.AV45"] <- "Diag.AV45_1" #rename one annoying column
colnames(wideformda)[colnames(wideformda)=="AV45_NONTP_CorticalSummary.CerebellarWhole_1"] <- "AV45_TP_CorticalSummary.CerebellarWhole_1"

colnames(wideformda)[colnames(wideformda)=="AV45_NONTP_CorticalSummary.BigRef_1"] <- "AV45_TP_CorticalSummary.BigRef_1"
```

```{r}
# vector of all variable names we want to include in our analyses (NEED TO ADD COLUMNS)
desired_column_names <- c("AV45_Date",
                          "Age.AV45",
                          "Diag.AV45",
                          "UW_MEM_AV45",
                          "UW_EF_AV45",
                          "AV45_TP_CorticalSummary.CerebellarWhole",
                          "AV45_TP_CorticalSummary.BigRef",
                          "Diagnosis_DATE")

# vector of variable names that don't change from visit to visit
baseline_vars <- c("SID","Gender","APOE2_BIN","APOE4_NUM","Edu..Yrs.")
```

```{r}
longformda <- list() # initialize empty list
# loop through each desired column above and create a list of data frames in long format with only the baseline variables and data from each visit for that column
for (colname in desired_column_names) {
  # all column beginning with the desired column name
  columns_to_reshape <- c(names(wideformda)[startsWith(names(wideformda), colname)])
  longformda[[colname]] <- reshape( # wide format data, baseline variables and desired columns only
    data=wideformda[c(baseline_vars,columns_to_reshape)],
    # specify that we're converting from wide to long format
    direction="long",
    # specify that we want to reshape these desired columns
    varying=columns_to_reshape,
    # define a new variable that will track what visit number the data comes from
    timevar="visitnumber",
    # indicate number of different visits with data for this variable
    times=c(as.character(1:length(columns_to_reshape))),
    # we name the column the desired column name
    v.names=colname,
    # indicate which variable we use to identify individuals
    idvar="SID")
}
```

```{r}
# now that we have a list of data frames, this code merges all the data frames together in a compact way
finalda <- Reduce(
  function(x,y,...) merge(x, y, by=c(baseline_vars,"visitnumber"), all=TRUE), # merge "by" all the baseline variables and visit number
  longformda)
# order data by subject ID and visit number
finalda$visitnumber <- as.integer(finalda$visitnumber)
finalda <- finalda[order(finalda$SID, finalda$visitnumber),]
finalda
```



# save formatted data frame to csv file for easy viewing
#write.csv(finalda, file.path(datapath,"longform_spreadsheet.csv"), row.names=FALSE)

```{r}
write.csv(finalda, file.path(datapath,"longform_spreadsheet.csv"), row.names=FALSE)
```
NOTES
summary <- c("Total", "Age at baseline, y","Female", "APOE4+", "Education, y", "Memory composite", "Executive function composite" )
total <- nrow(newdata)
mean_age <- mean(wideformda$Age.AV45_1)
values <- c(total, mean_age,  )
summary_table <- data.frame(summary, values)
summary_table


Defining our sample
```{r}
# subset to only cognitively normal individuals (Diag.AV45=="N")
newdata <- subset(finalda, Diag.AV45 =="N") 

# Goal: subset to only individuals with 2+ PET scans. Approach: subsetting to keep every row where SID is not NA for visitnumber 2

# subset to every SID that is not NA for visitnumber (!(is.na(visitnumber)))
newdata <- subset(newdata, !(is.na(visitnumber))) 
# create an empty vector (titled two_visit_SIDs)
two_visit_SIDs <- vector() 
# create a subset of newdata called second_visits where (visitnumber == 2)
second_visits <- subset(newdata, visitnumber == 2) 
# go through second_visits and add the SID from that row into two_visit_SIDs
for (i in 1:nrow(second_visits)) {
  current_SID = second_visits$SID[i]
  two_visit_SIDs <- c(two_visit_SIDs, current_SID)
}
# subset newdata to only include rows that contain SIDs in two_visit_SIDs (SID %in% two_visit_SIDs)
newdata <- subset(newdata, SID %in% two_visit_SIDs)

# compute summary statistics (see Table 1, don't need to include every single variable)
```

Computing other variables needed for analysis
```{r}
# compute baseline amyloid status (if AV45_NONTP_CorticalSummary/CerebellarWhole_1 > 1.11 (see paper), then AB+)

# compute "time" variable: amount of time (days, months, years, w/e) in between AV45 scan and cognitive test (use dates of each from spreadsheet)

# compute AV45 slope using linear regression (could be tricky, lmk if you need help:
# 1) run lm(AV45_TP_CorticalSummary... ~ Age) for each individual
# 2) the resulting "Estimate" value for Age is the AV45 slope for that individual
```

Spaghetti plots!
```{r}
# figure 1 from paper
# use ggplot to plot age vs. AV45 SUVR for AB-

# use ggplot to plot age vs. AV45 SUVR for AB+
```


Linear mixed effects model (LME)

```{r}
# look into nlme package, same basic structure as lm function: outcome ~ independent variables, also need to define:

# random effects (intercept and slope coded as "~1+time|SID")

# model control parameters (how hard your model tries to "converge", you can google this if interested but I would use these very liberal parameters: control = lmeControl(opt="optim",maxIter=100,msMaxIter=100,niterEM=50,msMaxEval=400))
```

Replicating Figure 3 from paper (if we have time)

```{r}
# if time and interesting in visualization
# ggplot to make scatter plot of memory slope vs. annualized AV45 change, fit with geom_smooth(method='lm')
```


