---
title: "ULAB_AD_Mina"
author: "Mina Baghai"
date: "3/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyr)
library(nlme)
library(ggplot2)
library(dplyr)
library(lubridate)
library(zoo)
```

``` {r}
# load data from csv into R dataframe
datapath <- file.path('/Users', 'minabaghai', 'OneDrive', 'Documents', 'Berkeley', 'ULab') # point to directory with data file
data_fname <- file.path(datapath, "Subjects_TP1.csv") # Path to spreadsheet 
wideformda <- read.csv(data_fname) # wide form since each visit has its own column, one row per individual
wideformda$baseline_AV45_TP_CorticalSummary.CerebellarWhole_1 <- wideformda$AV45_NONTP_CorticalSummary.CerebellarWhole_1
wideformda$baseline_AV45_TP_CorticalSummary.BigRef_1 <- wideformda$AV45_NONTP_CorticalSummary.BigRef_1
colnames(wideformda)[colnames(wideformda)=="Diag.AV45"] <- "Diag.AV45_1" #rename one annoying column
colnames(wideformda)[colnames(wideformda)=="AV45_NONTP_CorticalSummary.CerebellarWhole_1"] <- "AV45_TP_CorticalSummary.CerebellarWhole_1"

colnames(wideformda)[colnames(wideformda)=="AV45_NONTP_CorticalSummary.BigRef_1"] <- "AV45_TP_CorticalSummary.BigRef_1"

wideformda
```

```{r}
# vector of all variable names we want to include in our analyses (NEED TO ADD COLUMNS)
desired_column_names <- c("AV45_Date",
                          "Age.AV45",
                          "Diag.AV45",
                          "Diag.AV45_1",
                          "UW_MEM_AV45",
                          "UW_EF_AV45",
                          "AV45_TP_CorticalSummary.CerebellarWhole",
                          "baseline_AV45_TP_CorticalSummary.CerebellarWhole_1",
                          "AV45_TP_CorticalSummary.CerebellarWhole_2",
                          "AV45_TP_CorticalSummary.BigRef",
                          "baseline_AV45_TP_CorticalSummary.BigRef_1",
                          "AV45_TP_CorticalSummary.BigRef_2")
                          #"UW_TEST_EXAMDATE",
                          #"UW_EF",
                          #"UW_MEM")

# vector of variable names that don't change from visit to visit
baseline_vars <- c("SID","Gender","APOE2_BIN","APOE4_NUM","Edu..Yrs.")
```

```{r}
longformda <- list() # initialize empty list
# loop through each desired column above and create a list of data frames in long format with only the baseline variables and data from each visit for that column
for (colname in desired_column_names) {
  # all column beginning with the desired column name
  columns_to_reshape <- c(names(wideformda)[startsWith(names(wideformda), colname)])
  longformda[[colname]] <- reshape( # wide format data, baseline variables and desired columns only
    data=wideformda[c(baseline_vars,columns_to_reshape)],
    # specify that we're converting from wide to long format
    direction="long",
    # specify that we want to reshape these desired columns
    varying=columns_to_reshape,
    # define a new variable that will track what visit number the data comes from
    timevar="visitnumber",
    # indicate number of different visits with data for this variable
    times=c(as.character(1:length(columns_to_reshape))),
    # we name the column the desired column name
    v.names=colname,
    # indicate which variable we use to identify individuals
    idvar="SID")
}
```

```{r}
# now that we have a list of data frames, this code merges all the data frames together in a compact way
finalda <- Reduce(
  function(x,y,...) merge(x, y, by=c(baseline_vars,"visitnumber"), all=TRUE), # merge "by" all the baseline variables and visit number
  longformda)
# order data by subject ID and visit number
finalda$visitnumber <- as.integer(finalda$visitnumber)
finalda <- finalda[order(finalda$SID, finalda$visitnumber),]
finalda
```



# save formatted data frame to csv file for easy viewing
#write.csv(finalda, file.path(datapath,"longform_spreadsheet.csv"), row.names=FALSE)

```{r}
write.csv(finalda, file.path(datapath,"longform_spreadsheet.csv"), row.names=FALSE)
```

Defining our sample
```{r}
# subset to only cognitively normal individuals (Diag.AV45=="N") and to only individuals with 2+ PET scans (if AV45_TP_CorticalSummary/CerebellarWhole_2 is NA, remove)

newdata <- filter(finalda, Diag.AV45_1 == "N", AV45_TP_CorticalSummary.BigRef_2 != "NA", AV45_TP_CorticalSummary.CerebellarWhole_2 != "NA") 

newdata

nrow(newdata) #Gives us the number of cognitively normal subjects at baseline scan 241

finaldata <- inner_join(select(newdata, SID), finalda, by= "SID")
finaldata
```
Replicating Table 1 from the paper
```{r}
# compute summary statistics (see Table 1, don't need to include every single variable)

#Total (Value in Table 1: 142)
Total = length(unique(finaldata$SID))

#Florbetapir follow up time, y? (Value in Table 1: 3.9 ± 1.4)


#Age at baseline, y (Value in Table 1: 74.7 ± 7.0)
FirstVisit = filter(finaldata, visitnumber == 1) #filter for all first visits
AgeatBaseline = mean(FirstVisit$Age.AV45) #find mean age at first visit
AgeatBaselineSD = sd(FirstVisit$Age.AV45) #find st dev for age at first visit

#Female (Values in Table 1: 72 and 45%)
FilteredforFemale = filter(finaldata, Gender == 2) #filter for all females
Female = length(unique(FilteredforFemale$SID)) #find number of unique female SIDs
FemalePercent = Female / length(unique(finaldata$SID)) #number of females divided by number of participants

#APOE4+ (Values in Table 1: 32 and 15%)
APOE4Status = filter(finaldata, APOE4_NUM == 1) #filtered for all APOE4+
APOE4Plus = length(unique(APOE4Status$SID)) #find number of APOE4+
APOE4PlusPercent = APOE4Plus / length(unique(finaldata$SID)) #number of APOE4+ divided by number of participants

#Education, y (Value in Table 1: 16.7 ± 2.7)
Education = mean(FirstVisit$Edu..Yrs.) #find mean of education years at first visit
EducationSD = sd(FirstVisit$Edu..Yrs.) #find st dev for education at first visit

#Converted to MCI/AD (Values in Table 1: MCI = 12, AD = 2, Percent = 9.9%)
MCI = filter(finaldata, visitnumber != 1, Diag.AV45 == "LMCI") #find anyone MCI at any point
ConvertedtoMCI = length(unique(MCI$SID)) #number of people who converted to MCI because had filtered to cognitively normal at baseline

AD = filter(finaldata, visitnumber != 1, Diag.AV45 == "AD") #find anyone AD at any point
ConvertedtoAD = length(unique(AD$SID)) #number of people who converted to AD because had filtered to cognitively normal at baseline

ConvertedtoMCIorADPercent = (ConvertedtoMCI + ConvertedtoAD)/length(unique(finaldata$SID)) #number of people converted to MCI or AD divided by number of participants

#Converted to florbetapir+ (Values in table 1: 13 and 9.2%)


#Memory Composite (Value in Table 1: 1.12 ± 0.58)
Memory = mean(FirstVisit$UW_MEM_AV45) #find mean of memory score at first visit
MemorySD = sd(FirstVisit$UW_MEM_AV45) #find st dev for memory score at first visit

#Executive Function Compostie (Value in Table 1: 1.00 ± 0.67))
ExecFunction = mean(FirstVisit$UW_EF_AV45) #find mean of exec function score at first visit
ExecFunctionSD = sd(FirstVisit$UW_EF_AV45) #find st dev for exec function score at first visit

#Cognitive Testing Follow-up, y (Value in Table 1: 3.8 ± 1.0)

```

Computing other variables needed for analysis (baseline amyloid status)
```{r}
# compute baseline amyloid status (if AV45_NONTP_CorticalSummary/CerebellarWhole_1 > 1.11 (see paper), then AB+)
baseline_amyloid_status <- c()

for (i in 1:length(newdata$SID))
  if (newdata$baseline_AV45_TP_CorticalSummary.BigRef_1[i] < 0.79 &&
      newdata$baseline_AV45_TP_CorticalSummary.CerebellarWhole_1[i] < 1.11) {
      baseline_amyloid_status <- baseline_amyloid_status %>% append("AB-")
  }else if (newdata$baseline_AV45_TP_CorticalSummary.BigRef_1[i] > 0.79 &&
      newdata$baseline_AV45_TP_CorticalSummary.CerebellarWhole_1[i] > 1.11) {
      baseline_amyloid_status <- baseline_amyloid_status %>% append("AB+")
  } else {
      baseline_amyloid_status <- baseline_amyloid_status %>% append("discordant") 
  }

baseline_amyloid_status
table(baseline_amyloid_status) #241 participants, 157 AB-, 61 AB+, 23 discordant 

newdata$baseline_amyloid_status <- baseline_amyloid_status

finaldata <- inner_join(select(newdata, SID, baseline_amyloid_status), finalda, by= "SID")
finaldata

```

Computing the time variable
```{r}
# compute "time" variable: amount of time (days, months, years, w/e) in between AV45 scan and cognitive test (use dates of each from spreadsheet)
finaldata$AV45_Date <- mdy(finaldata$AV45_Date) # convert date strings to datetime using lubridate package
finaldata$time <- as.data.frame(finaldata %>% group_by(SID) %>% mutate(time = (AV45_Date - AV45_Date[1])/365))$time # create new variable "time" in years between first AV45 visit and current visit

finaldata
```
Compute AV45 Slope
```{r}
# compute AV45 slope using linear regression (could be tricky, lmk if you need help:
# 1) run lm(AV45_TP_CorticalSummary... ~ Age) for each individual
# 2) the resulting "Estimate" value for Age is the AV45 slope for that individual

finaldata$annualized_AV45 <- as.data.frame(finaldata %>% group_by(SID) %>% # for each individual
  # regress AV45 SUVR against time (more accurate than Age), extract estimate for time (slope)
  mutate(annualized_AV45 = lm(AV45_TP_CorticalSummary.BigRef ~ time)$coefficients["time"]))$annualized_AV45

finaldata$annualized_AV45
finaldata
```

Spaghetti plots!
```{r}
# figure 1 from paper
# use ggplot to plot age vs. AV45 SUVR for AB-

baseline_negative <- finaldata %>% select(SID, Age.AV45, AV45_TP_CorticalSummary.BigRef, baseline_amyloid_status) %>% filter(baseline_amyloid_status == "AB-") %>% na.omit

baseline_negative$SID <- as.factor(baseline_negative$SID) #We want SID to be categorical variable, otherwise R thinks it's a numerical value and will assign colors based on gradient

p <- ggplot(data = baseline_negative, aes(x = Age.AV45, y = AV45_TP_CorticalSummary.BigRef, group = SID)) +geom_point(aes(color = SID)) + geom_line(aes(color = SID)) + geom_hline(yintercept = 0.79, linetype = "dotted") + theme(legend.position = "none") +xlim(50, 100)+ylim(0.65, 0.95) + ggtitle("Baseline Negative") + labs(y="Florbetapir SUVR", x = "Age (years)")+
  theme(plot.title = element_text(hjust = 0.5))

p

# use ggplot to plot age vs. AV45 SUVR for AB+

baseline_positive <- finaldata %>% select(SID, Age.AV45, AV45_TP_CorticalSummary.BigRef, baseline_amyloid_status) %>% filter(baseline_amyloid_status == "AB+") %>% na.omit

baseline_positive$SID <- as.factor(baseline_positive$SID) #We want SID to be categorical variable, otherwise R thinks it's a numerical value and will assign colors based on gradient

p2 <- ggplot(data = baseline_positive, aes(x = Age.AV45, y = AV45_TP_CorticalSummary.BigRef, group = SID)) +geom_point(aes(color = SID)) + geom_line(aes(color = SID)) + geom_hline(yintercept = 0.79, linetype = "dotted") + theme(legend.position = "none") +xlim(50, 100)+ylim(0.70, 1.25) + ggtitle("Baseline Positive") + labs(y="Florbetapir SUVR", x = "Age (years)")+
  theme(plot.title = element_text(hjust = 0.5))

p2
```

Linear mixed effects model (LME)
```{r}
# look into nlme package, same basic structure as lm function: outcome ~ independent variables, also need to define:

# random effects (intercept and slope coded as "~1+time|SID")

# model control parameters (how hard your model tries to "converge", you can google this if interested but I would use these very liberal parameters: control = lmeControl(opt="optim",maxIter=100,msMaxIter=100,niterEM=50,msMaxEval=400))

memory_lme <- lme((UW_MEM_AV45 ~ Gender + APOE4_NUM + Edu..Yrs. + Age.AV45 + baseline_AV45_TP_CorticalSummary.BigRef_1 + time + baseline_AV45_TP_CorticalSummary.BigRef_1 + annualized_AV45 + time * annualized_AV45), data = na.locf(finaldata), random = ~1+time|SID )

ef_lme <- lme((UW_EF_AV45 ~ Gender + APOE4_NUM + Edu..Yrs. + Age.AV45 + baseline_AV45_TP_CorticalSummary.BigRef_1 + time + baseline_AV45_TP_CorticalSummary.BigRef_1 + annualized_AV45 + time * annualized_AV45), data = na.locf(finaldata), random = ~1+time|SID )

```

Replicating Figure 3 from paper
```{r}
# if time and interesting in visualization
# ggplot to make scatter plot of memory slope vs. annualized AV45 change, fit with geom_smooth(method='lm')
```


